---
title: "Zero-Inflated Models: A Comprehensive Tutorial"
description: "A complete guide to understanding and implementing zero-inflated models for count data analysis"
author: "Andrea Di Francia"
date: "2025-01-09"
categories: [statistics, modeling, R, tutorial]
editor: 
   render-on-save: true
format:
  html:
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    code-copy: true
    code-tools: true
    code-line-numbers: true
    code-overflow: wrap
    fig-width: 8
    fig-height: 5
    fig-dpi: 144
    fig-align: center
    fig-cap-location: bottom
    tbl-cap-location: top
    anchor-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
mermaid:
  theme: default
  themeVariables:
    primaryColor: "#0d6efd"
    primaryTextColor: "#ffffff"
    primaryBorderColor: "#0d6efd"
    lineColor: "#0d6efd"
    secondaryColor: "#e9ecef"
    tertiaryColor: "#f8f9fa"
    fontFamily: "Inter, system-ui, -apple-system, Segoe UI, Roboto, Helvetica, Arial, \"Apple Color Emoji\", \"Segoe UI Emoji\""
---

# Introduction

Zero-inflated models represent a crucial statistical framework for analyzing count data that contains an excess of zero observations beyond what standard distributions would predict. This tutorial provides a comprehensive guide to understanding the theoretical foundations, practical implementation, and advanced applications of zero-inflated models in R.

## Learning Objectives

By the end of this tutorial, you will be able to:

- Understand the mathematical foundations of generalized linear models and the exponential family
- Recognize when zero-inflated models are needed for your data
- Implement zero-inflated Poisson and negative binomial models using `glmmTMB`
- Perform comprehensive model diagnostics using `DHARMa` and `performance` packages
- Distinguish between zero-inflated and hurdle models
- Work with advanced count distributions for complex data structures

## When Do We Need Zero-Inflated Models?

Zero-inflated models address a fundamental challenge in count data analysis: **excess zeros**. Consider these real-world scenarios:

- **Healthcare utilization**: Some patients cannot access healthcare (structural zeros), while others simply don't need it during the study period (sampling zeros)
- **Species abundance**: Some sites may be unsuitable for a species (structural zeros), while suitable sites might not show the species during sampling (sampling zeros)
- **Patent applications**: Some companies never file patents (structural zeros), while others may not file in a given year (sampling zeros)

```{mermaid}
flowchart TD
    A[Count Data with Excess Zeros] --> B{Can zeros arise from two distinct processes?}
    B -->|Yes| C[Zero-Inflated Models]
    B -->|No| D[Standard Count Models]
    C --> E[Structural Zeros: Cannot experience event]
    C --> F[Sampling Zeros: Could experience event but didn't]
    D --> G[Poisson/Negative Binomial]
```

# Theoretical Foundation: Generalized Linear Models

## The Exponential Family

The foundation of zero-inflated models rests upon the **exponential family** of distributions. A distribution belongs to the exponential family if its probability density can be written as:

$$f(y|\theta, \phi) = \exp\left\{\frac{y\theta - b(\theta)}{\phi} + c(y, \phi)\right\}$$

where:
- $\theta$ is the **natural parameter** 
- $\phi$ is the **dispersion parameter**
- $b(\cdot)$ and $c(\cdot)$ are specific functions that define the distribution

### Key Examples

**Poisson Distribution:**
$$f(y|\mu) = \frac{\mu^y e^{-\mu}}{y!} = \exp\{y \log \mu - \mu - \log y!\}$$

This gives us:
- Natural parameter: $\theta = \log \mu$
- Cumulant function: $b(\theta) = e^\theta = \mu$
- Canonical link: $\log(\mu)$ (log link)

**Bernoulli Distribution:**
$$f(y|\mu) = \mu^y(1-\mu)^{1-y} = \exp\left\{y \log\left(\frac{\mu}{1-\mu}\right) + \log(1-\mu)\right\}$$

This gives us:
- Natural parameter: $\theta = \log\left(\frac{\mu}{1-\mu}\right)$ (logit)
- Canonical link: logit link

## Generalized Linear Models (GLMs)

GLMs extend linear regression through three components:

1. **Linear predictor**: $\eta = X\beta$ 
2. **Link function**: $g(\mu) = \eta$
3. **Response distribution**: From exponential family

```{mermaid}
graph LR
    A[Predictors X] --> B[Linear Predictor η = Xβ]
    B --> C[Link Function g(μ) = η]
    C --> D[Expected Response μ]
    D --> E[Response Distribution Y]
```

### Logistic Regression Example

For binary outcomes, we use:
- **Distribution**: Bernoulli 
- **Link function**: Logit link $\log\left(\frac{\pi}{1-\pi}\right) = X\beta$
- **Interpretation**: $\exp(\beta_j)$ = odds ratio for unit change in $x_j$

### Poisson Regression Example

For count data, we use:
- **Distribution**: Poisson
- **Link function**: Log link $\log(\lambda) = X\beta$  
- **Interpretation**: $\exp(\beta_j)$ = multiplicative effect on rate for unit change in $x_j$

**Key assumption**: $E[Y] = \text{Var}(Y) = \lambda$ (equidispersion)

# Practical Implementation: Foundation Models

## Package Installation

Before starting, install the required packages:

```{r install, eval=FALSE}
# Install packages if not already available
install.packages(c("glmmTMB", "DHARMa", "performance", "MASS", 
                   "ggplot2", "dplyr", "gridExtra", "knitr", "caret"))
```

Let's start with practical examples in R. First, we'll set up our environment and create a realistic dataset:

```{r setup, message=FALSE, warning=FALSE, eval=FALSE}
# Load required packages
library(glmmTMB)
library(DHARMa)
library(performance)
library(MASS)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(knitr)

# Set seed for reproducibility
set.seed(123)

# Create realistic fishing dataset
n <- 500
fishing_data <- data.frame(
  group_size = sample(1:6, n, replace = TRUE),
  hours_fishing = runif(n, 0, 8),
  experience = sample(c("novice", "intermediate", "expert"), n, replace = TRUE),
  bait_quality = rnorm(n, 0, 1),
  weather_good = rbinom(n, 1, 0.7)
)

# Create zero-inflated fish counts
# Some people don't fish at all (structural zeros)
prob_not_fishing <- plogis(-1 + 0.5 * (fishing_data$experience == "novice") - 
                          0.3 * fishing_data$weather_good)
not_fishing <- rbinom(n, 1, prob_not_fishing)

# For those who fish, generate counts from negative binomial
lambda <- exp(0.5 + 0.2 * fishing_data$group_size + 
              0.3 * fishing_data$hours_fishing +
              0.4 * (fishing_data$experience == "expert") +
              0.2 * fishing_data$bait_quality +
              0.3 * fishing_data$weather_good)

fish_counts <- ifelse(not_fishing == 1, 0, 
                     rnbinom(n, size = 2, mu = lambda))

fishing_data$fish_caught <- fish_counts
fishing_data$experience <- factor(fishing_data$experience, 
                                 levels = c("novice", "intermediate", "expert"))

# Display basic statistics
cat("Proportion of zeros:", round(mean(fishing_data$fish_caught == 0), 3), "\n")
cat("Mean fish count:", round(mean(fishing_data$fish_caught), 2), "\n")
cat("Variance:", round(var(fishing_data$fish_caught), 2), "\n")
cat("Variance-to-mean ratio:", round(var(fishing_data$fish_caught)/mean(fishing_data$fish_caught), 2))
```

## Data Exploration

```{r exploration, message=FALSE, warning=FALSE, eval=FALSE}
# Visualize the distribution
p1 <- ggplot(fishing_data, aes(x = fish_caught)) +
  geom_histogram(binwidth = 1, fill = "skyblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Fish Caught", x = "Fish Count", y = "Frequency") +
  theme_minimal()

# Check zero inflation by groups
p2 <- fishing_data %>%
  group_by(experience) %>%
  summarise(prop_zeros = mean(fish_caught == 0),
            mean_count = mean(fish_caught), .groups = 'drop') %>%
  ggplot(aes(x = experience, y = prop_zeros)) +
  geom_col(fill = "coral", alpha = 0.7, color = "black") +
  labs(title = "Proportion of Zeros by Experience", 
       x = "Experience Level", y = "Proportion of Zeros") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

The high proportion of zeros (>40%) and variance-to-mean ratio >>1 suggest we need zero-inflated models.

## Foundation Models

### 1. Logistic Regression for Zero vs Non-Zero

```{r logistic, eval=FALSE}
# Model the probability of catching any fish
zero_indicator <- ifelse(fishing_data$fish_caught == 0, 1, 0)
logistic_model <- glm(zero_indicator ~ group_size + hours_fishing + experience + 
                     bait_quality + weather_good, 
                     data = fishing_data, family = binomial())

summary(logistic_model)
```

**Interpretation**: 
- Negative coefficients indicate factors that **decrease** the probability of zero catches
- `hours_fishing` has coefficient ≈ -0.5, so each additional hour reduces the log-odds of zero catch by 0.5
- Experienced anglers have much lower probability of zero catches

### 2. Poisson Regression for Count Data

```{r poisson, eval=FALSE}
# Standard Poisson model (ignoring zero-inflation)
poisson_model <- glm(fish_caught ~ group_size + hours_fishing + experience + 
                    bait_quality + weather_good, 
                    data = fishing_data, family = poisson())

summary(poisson_model)

# Check for overdispersion
overdispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) / 
                       df.residual(poisson_model)
cat("Overdispersion ratio:", round(overdispersion_ratio, 2))
```

The overdispersion ratio is much greater than 1, indicating severe overdispersion, violating the Poisson assumption that variance equals the mean.

# Zero-Inflated Models: Theory and Implementation

## Mathematical Formulation

### Zero-Inflated Poisson (ZIP) Model

The ZIP model assumes that zeros can arise from two sources:

$$P(Y = 0) = \pi + (1-\pi)e^{-\lambda}$$
$$P(Y = k) = (1-\pi)\frac{\lambda^k e^{-\lambda}}{k!}, \quad k = 1, 2, 3, ...$$

where:
- $\pi \in [0,1]$ = probability of structural zero (non-participants)
- $\lambda > 0$ = Poisson rate for participants
- $(1-\pi)$ = probability of being a participant

**Mean and Variance:**
$$E[Y] = (1-\pi)\lambda$$
$$\text{Var}(Y) = \lambda(1-\pi)(1 + \pi\lambda)$$

Note that $\text{Var}(Y) > E[Y]$ always, showing natural overdispersion.

### Zero-Inflated Negative Binomial (ZINB) Model

The ZINB extends ZIP to handle additional overdispersion:

$$P(Y = 0) = \pi + (1-\pi)\left(\frac{r}{r+\mu}\right)^r$$
$$P(Y = k) = (1-\pi)\frac{\Gamma(k+r)}{k!\Gamma(r)}\left(\frac{r}{r+\mu}\right)^r\left(\frac{\mu}{r+\mu}\right)^k$$

where $\mu$ is the mean parameter and $r$ controls dispersion.

## Implementation with glmmTMB

```{r zero_inflated_models, eval=FALSE}
# Zero-Inflated Poisson (ZIP) model
zip_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                     bait_quality + weather_good,
                     ziformula = ~ experience + weather_good,
                     data = fishing_data,
                     family = poisson())

# Zero-Inflated Negative Binomial (ZINB) model  
zinb_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                      bait_quality + weather_good,
                      ziformula = ~ experience + weather_good,
                      data = fishing_data,
                      family = nbinom2())

# Model summaries
cat("=== Zero-Inflated Poisson Model ===\n")
summary(zip_model)

cat("\n=== Zero-Inflated Negative Binomial Model ===\n") 
summary(zinb_model)
```

## Interpreting Results

```{r interpretation, eval=FALSE}
# Extract coefficients from ZINB model
count_coefs <- fixef(zinb_model)$cond
zi_coefs <- fixef(zinb_model)$zi

cat("Count model coefficients (log scale):\n")
print(round(count_coefs, 3))

cat("\nZero-inflation coefficients (logit scale):\n")
print(round(zi_coefs, 3))

# Convert to interpretable scales
cat("\n=== INTERPRETATIONS ===\n")
cat("Count model - Multiplicative effects on expected catch:\n")
multiplicative_effects <- exp(count_coefs)
for(i in 1:length(multiplicative_effects)) {
  effect <- multiplicative_effects[i]
  name <- names(multiplicative_effects)[i]
  if(name == "(Intercept)") {
    cat("Baseline expected catch:", round(effect, 2), "fish\n")
  } else if(grepl("experience", name)) {
    cat(name, ": ", round((effect-1)*100, 1), "% change in expected catch\n", sep="")
  } else {
    cat(name, ": ", round((effect-1)*100, 1), "% change per unit increase\n", sep="")
  }
}

cat("\nZero-inflation - Probability of being a non-fisher:\n")
zi_probs <- plogis(zi_coefs)
for(i in 1:length(zi_probs)) {
  cat(names(zi_probs)[i], ":", round(zi_probs[i], 3), "\n")
}
```

**Key Insights:**
- **Count Model**: Shows how factors affect the expected catch for those who do fish
- **Zero-Inflation Model**: Shows how factors affect the probability of not fishing at all
- Expert anglers have both higher catch rates AND lower probability of being non-fishers

# Model Diagnostics with DHARMa and Performance

Proper diagnostics are crucial for zero-inflated models due to their complex structure.

```{r diagnostics, fig.height=8, eval=FALSE}
# DHARMa diagnostics for ZINB model
residuals_sim <- simulateResiduals(zinb_model, plot = FALSE)
plot(residuals_sim, main = "DHARMa Residuals for ZINB Model")

# Specific diagnostic tests
cat("=== DHARMa Diagnostic Tests ===\n")
zero_test <- testZeroInflation(residuals_sim)
cat("Zero-inflation test p-value:", round(zero_test$p.value, 4), "\n")

disp_test <- testDispersion(residuals_sim)  
cat("Dispersion test p-value:", round(disp_test$p.value, 4), "\n")

outlier_test <- testOutliers(residuals_sim)
cat("Outliers test p-value:", round(outlier_test$p.value, 4), "\n")
```

```{r performance_checks, eval=FALSE}
# Performance package diagnostics
cat("=== Performance Package Diagnostics ===\n")
model_perf <- model_performance(zinb_model)
print(model_perf)

# Check specific issues
cat("\nOverdispersion check:\n")
overdisp_check <- check_overdispersion(zinb_model)
print(overdisp_check)

cat("\nZero-inflation check:\n")
zi_check <- check_zeroinflation(zinb_model)
print(zi_check)
```

## Model Comparison

```{r model_comparison, eval=FALSE}
# Compare different model approaches
models <- list(
  "Poisson" = poisson_model,
  "ZIP" = zip_model, 
  "ZINB" = zinb_model
)

# Extract model comparison metrics
comparison_table <- compare_performance(models, metrics = c("AIC", "BIC", "RMSE"))
print(comparison_table)

# Likelihood ratio tests for nested models
cat("\n=== Likelihood Ratio Tests ===\n")
# Test ZIP vs Poisson
lrt_zip <- anova(zip_model, poisson_model)
cat("ZIP vs Poisson: p-value =", round(lrt_zip$`Pr(>Chisq)`[2], 4), "\n")

# Test ZINB vs ZIP  
lrt_zinb <- anova(zinb_model, zip_model)
cat("ZINB vs ZIP: p-value =", round(lrt_zinb$`Pr(>Chisq)`[2], 4), "\n")
```

The ZINB model wins on all criteria, confirming our need for both zero-inflation and overdispersion modeling.

# Advanced Topic: Hurdle Models

## Conceptual Difference from Zero-Inflated Models

**Zero-Inflated Models**: Zeros arise from a **mixture** of two processes
- Some zeros are "structural" (never could have non-zero)  
- Some zeros are "sampling" (could have been non-zero but weren't)

**Hurdle Models**: **All** zeros arise from one process, positive values from another
- First process: Will there be any events? (Binary)
- Second process: How many events? (Truncated count)

```{mermaid}
graph TD
    A[Observed Zero] --> B{Model Type}
    B --> C[Zero-Inflated: Could be structural OR sampling zero]
    B --> D[Hurdle: Single process determines all zeros]
    
    E[Observed Positive Count] --> F{Model Type}  
    F --> G[Zero-Inflated: Standard count distribution]
    F --> H[Hurdle: Truncated count distribution]
```

## Mathematical Formulation

**Hurdle Model:**
$$P(Y = 0) = \theta$$
$$P(Y = k) = (1-\theta) \times \frac{f_{\text{truncated}}(k|\lambda)}{1-P_{\text{count}}(0)}, \quad k = 1, 2, 3, ...$$

**Key difference**: Hurdle models can handle **zero deflation** (fewer zeros than expected), while zero-inflated models cannot.

## Implementation

```{r hurdle_models, eval=FALSE}
# Hurdle Poisson model
hurdle_pois <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                       bait_quality + weather_good,
                       ziformula = ~ experience + weather_good,
                       data = fishing_data,
                       family = truncated_poisson())

# Hurdle Negative Binomial model
hurdle_nb <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                     bait_quality + weather_good,
                     ziformula = ~ experience + weather_good,
                     data = fishing_data,
                     family = truncated_nbinom2())

# Compare all models
all_models <- list(
  "Poisson" = poisson_model,
  "ZIP" = zip_model,
  "ZINB" = zinb_model, 
  "Hurdle_Poisson" = hurdle_pois,
  "Hurdle_NB" = hurdle_nb
)

comparison_all <- compare_performance(all_models, metrics = c("AIC", "BIC"))
print(comparison_all)
```

## When to Choose Hurdle vs Zero-Inflated?

**Choose Zero-Inflated when:**
- Strong theoretical justification for two distinct zero-generating processes
- Zeros are genuinely "extra" beyond what the count distribution predicts
- You want to model factors affecting structural vs sampling zeros differently

**Choose Hurdle when:**
- All zeros conceptually arise from the same process  
- You need flexibility to handle zero deflation
- You want to model "participation" vs "intensity" separately

# Advanced Count Distributions

## Alternative Distributions for Count Data

### 1. Negative Binomial Parameterizations

```{r nb_parameterizations, eval=FALSE}
# NB1: Variance = μ(1 + α) [linear variance]
zinb1_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                       bait_quality + weather_good,
                       ziformula = ~ experience + weather_good,
                       data = fishing_data,
                       family = nbinom1())

# NB2: Variance = μ(1 + μ/k) [quadratic variance] 
zinb2_model <- zinb_model  # This is our original model

cat("NB1 vs NB2 comparison:\n")
nb_comparison <- compare_performance(zinb1_model, zinb2_model)
print(nb_comparison)
```

**When to use NB1 vs NB2:**
- **NB1**: When variance increases linearly with mean
- **NB2**: When variance increases quadratically with mean (more common)

### 2. Generalized Poisson for Under/Overdispersion

```{r generalized_poisson, eval=FALSE}
# Generalized Poisson can handle both under and overdispersion
zigp_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                      bait_quality + weather_good,
                      ziformula = ~ experience + weather_good,
                      data = fishing_data,
                      family = genpois())

summary(zigp_model)
```

### 3. Conway-Maxwell-Poisson for Flexible Dispersion

```{r conway_maxwell, eval=FALSE}
# Conway-Maxwell-Poisson provides very flexible dispersion modeling
# Note: Requires additional computational resources
zicomp_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                        bait_quality + weather_good,
                        ziformula = ~ experience + weather_good,
                        data = fishing_data,
                        family = compois())
```

### 4. Zero-Altered Models for Continuous Data

For continuous data with zeros (e.g., biomass, rainfall):

```{r zero_altered_gamma, eval=FALSE}
# Example: Zero-Altered Gamma for biomass data
# Create example biomass data
biomass_data <- data.frame(
  treatment = factor(rep(c("control", "fertilized"), each = 200)),
  soil_moisture = rnorm(400, 50, 15),
  temperature = rnorm(400, 25, 5)
)

# Generate zero-altered gamma data
prob_zero <- plogis(-2 + 0.5 * (biomass_data$treatment == "control"))
has_biomass <- rbinom(400, 1, 1 - prob_zero)
gamma_rate <- exp(-1 + 0.3 * (biomass_data$treatment == "fertilized"))
biomass_data$biomass <- ifelse(has_biomass == 0, 0, 
                              rgamma(400, shape = 2, rate = gamma_rate))

# Zero-Altered Gamma model
zag_model <- glmmTMB(biomass ~ treatment + soil_moisture + temperature,
                     ziformula = ~ treatment + soil_moisture,
                     data = biomass_data,
                     family = Gamma(link = "log"))
```

# Model Selection and Validation

## Cross-Validation for Robust Model Selection

```{r cross_validation, eval=FALSE}
library(caret)

# Create folds for cross-validation
set.seed(123)
folds <- createFolds(fishing_data$fish_caught, k = 5)

# Function to calculate RMSE
calculate_rmse <- function(model, test_data) {
  pred <- predict(model, newdata = test_data, type = "response")
  sqrt(mean((test_data$fish_caught - pred)^2))
}

# Cross-validation for key models
cv_results <- data.frame(
  Fold = 1:5,
  ZIP_RMSE = numeric(5),
  ZINB_RMSE = numeric(5),
  Hurdle_NB_RMSE = numeric(5)
)

for(i in 1:5) {
  train_data <- fishing_data[-folds[[i]], ]
  test_data <- fishing_data[folds[[i]], ]
  
  # Fit models on training data
  zip_cv <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                    bait_quality + weather_good,
                    ziformula = ~ experience + weather_good,
                    data = train_data, family = poisson())
  
  zinb_cv <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                     bait_quality + weather_good,
                     ziformula = ~ experience + weather_good,
                     data = train_data, family = nbinom2())
  
  hurdle_cv <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                       bait_quality + weather_good,
                       ziformula = ~ experience + weather_good,
                       data = train_data, family = truncated_nbinom2())
  
  # Calculate RMSE on test data
  cv_results$ZIP_RMSE[i] <- calculate_rmse(zip_cv, test_data)
  cv_results$ZINB_RMSE[i] <- calculate_rmse(zinb_cv, test_data)
  cv_results$Hurdle_NB_RMSE[i] <- calculate_rmse(hurdle_cv, test_data)
}

# Summarize cross-validation results
cv_summary <- cv_results %>%
  summarise(
    ZIP_Mean = mean(ZIP_RMSE),
    ZIP_SD = sd(ZIP_RMSE),
    ZINB_Mean = mean(ZINB_RMSE),
    ZINB_SD = sd(ZINB_RMSE),
    Hurdle_Mean = mean(Hurdle_NB_RMSE),
    Hurdle_SD = sd(Hurdle_NB_RMSE)
  )

print(cv_summary)
```

## Final Model Selection Framework

```{r final_selection, eval=FALSE}
# Comprehensive model selection criteria
final_selection <- function(models_list, cv_results) {
  criteria <- data.frame(
    Model = names(models_list),
    AIC = sapply(models_list, AIC),
    BIC = sapply(models_list, BIC),
    stringsAsFactors = FALSE
  )
  
  # Add cross-validation RMSE
  criteria$CV_RMSE <- c(
    mean(cv_results$ZIP_RMSE),
    mean(cv_results$ZINB_RMSE), 
    mean(cv_results$Hurdle_NB_RMSE)
  )
  
  # Rank models (lower is better)
  criteria$AIC_rank <- rank(criteria$AIC)
  criteria$BIC_rank <- rank(criteria$BIC)
  criteria$CV_rank <- rank(criteria$CV_RMSE)
  criteria$Average_rank <- rowMeans(criteria[, c("AIC_rank", "BIC_rank", "CV_rank")])
  
  return(criteria[order(criteria$Average_rank), ])
}

# Apply selection framework
key_models <- list(
  "ZIP" = zip_model,
  "ZINB" = zinb_model,
  "Hurdle_NB" = hurdle_nb
)

selection_results <- final_selection(key_models, cv_results)
print(selection_results)

best_model_name <- selection_results$Model[1]
cat("\n*** WINNER:", best_model_name, "***\n")
```

# Predictions and Interpretation

## Making Predictions from Zero-Inflated Models

```{r predictions, eval=FALSE}
# Create prediction scenarios
new_scenarios <- data.frame(
  group_size = c(2, 4, 2, 4),
  hours_fishing = c(2, 6, 2, 6),
  experience = factor(c("novice", "novice", "expert", "expert"), 
                     levels = c("novice", "intermediate", "expert")),
  bait_quality = c(0, 0, 1, 1),
  weather_good = c(1, 1, 1, 1)
)

# Get our best model (ZINB)
final_model <- zinb_model

# Different types of predictions
predictions <- data.frame(
  new_scenarios,
  expected_catch = predict(final_model, newdata = new_scenarios, type = "response"),
  prob_zero_inflation = predict(final_model, newdata = new_scenarios, type = "zprob"),
  conditional_catch = predict(final_model, newdata = new_scenarios, type = "conditional")
)

predictions$prob_any_fish <- 1 - predictions$prob_zero_inflation

print(predictions)
```

**Interpretation:**
- `expected_catch`: Overall expected number of fish (includes zero-inflation)
- `prob_zero_inflation`: Probability of being a "structural zero" (non-fisher)
- `conditional_catch`: Expected catch IF the person fishes
- `prob_any_fish`: Probability of catching at least one fish

## Marginal Effects Visualization

```{r marginal_effects, eval=FALSE}
# Create effect data for hours fishing
effect_data <- expand.grid(
  group_size = 3,  # Average group size
  hours_fishing = seq(0, 8, 0.5),
  experience = "intermediate",
  bait_quality = 0,
  weather_good = 1
)

# Calculate predictions with confidence intervals
effect_predictions <- predict(final_model, newdata = effect_data, 
                             type = "response", se.fit = TRUE)

effect_plot_data <- data.frame(
  effect_data,
  predicted = effect_predictions$fit,
  lower = effect_predictions$fit - 1.96 * effect_predictions$se.fit,
  upper = effect_predictions$fit + 1.96 * effect_predictions$se.fit
)

# Plot marginal effect
ggplot(effect_plot_data, aes(x = hours_fishing, y = predicted)) +
  geom_line(color = "blue", size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  labs(title = "Expected Fish Catch by Hours Fishing",
       subtitle = "For intermediate angler, group size 3, good weather",
       x = "Hours Fishing", 
       y = "Expected Fish Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

# Summary and Best Practices

## Key Takeaways

1. **Zero-inflated models are essential** when you have excess zeros from two distinct processes
2. **ZINB often outperforms ZIP** due to its ability to handle additional overdispersion  
3. **Hurdle models offer more flexibility** and can handle zero deflation
4. **Model diagnostics are crucial** - use DHARMa and performance packages
5. **Cross-validation provides robust model selection** beyond just AIC/BIC

## Best Practices Checklist

✅ **Before modeling:**
- Explore your data for excess zeros
- Consider the data generating process theoretically
- Check for overdispersion in standard Poisson models

✅ **During modeling:**
- Compare multiple model types (ZIP, ZINB, Hurdle)
- Use different covariates for count and zero-inflation components when justified
- Check convergence and model diagnostics

✅ **After modeling:**
- Validate with cross-validation or holdout data
- Interpret both components of the model
- Make predictions appropriate to your research questions

## When to Use Each Model Type

| Model Type | Best When | Avoid When |
|------------|-----------|------------|
| **Standard Poisson** | No excess zeros, equidispersed | Overdispersed or zero-inflated |
| **Negative Binomial** | Overdispersed but no excess zeros | Clear zero-inflation |
| **Zero-Inflated Poisson** | Excess zeros, minimal overdispersion | Heavy overdispersion |
| **Zero-Inflated NB** | Excess zeros AND overdispersion | Zero deflation |
| **Hurdle Models** | Want flexibility, possible zero deflation | Strong theory for mixture processes |

## Further Reading

- Zeileis, A., Kleiber, C., & Jackman, S. (2008). Regression models for count data in R. *Journal of Statistical Software*, 27(8), 1-25.
- Lambert, D. (1992). Zero-inflated Poisson regression, with an application to defects in manufacturing. *Technometrics*, 34(1), 1-14.
- Mullahy, J. (1986). Specification and testing of some modified count data models. *Journal of Econometrics*, 33(3), 341-365.

---

*This tutorial provides a comprehensive foundation for zero-inflated modeling. Remember that the choice of model should always be guided by both statistical evidence and theoretical understanding of your data generating process.*
