---
title: "Zero-Inflated Models: A Comprehensive Tutorial"
description: "A practical guide to understanding and implementing zero-inflated models for count data analysis"
author: "Andrea Di Francia"
date: "2026-06-15"
categories: [statistics, modeling, R, tutorial]
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc-depth: 3
    number-sections: true
    code-copy: true
    code-overflow: wrap
    fig-align: center
    fig-cap-location: bottom
    tbl-cap-location: top
---

# Core Topics

## Introduction

Testing

### What Are Zero-Inflation Models?

In regression analysis, zero-inflated models are a family of statistical methods used to analyze count data when there are more zeros than simple count models (like Poisson) expect. In plain terms: when you're counting events (visits to a clinic, number of fish caught, defects on a product) and a lot of observations are zero, standard models can give misleading results. Zero-inflated models help by explicitly accounting for two ways a zero can occur:

- one "structural" (the event cannot happen for some units)
- one "sampling" (the event could happen but didn't in that observation).

### When Do We Need Zero-Inflated Models?

Zero-inflated models address a fundamental challenge in count data analysis: **excess zeros**. Consider these real-world scenarios:

 - **Healthcare utilization (improved)**: Some patients are effectively unable to generate events because they never access services (structural zeros) — for example, people living in remote areas with no nearby clinics, or those excluded by policy or eligibility criteria. Other patients could use services but happen not to during the study window (sampling zeros) — for example, healthy patients who simply don't visit in that period or patients whose condition didn't require care.
 - **Disease surveillance (infection detection)**: Some locations truly have zero prevalence for a pathogen because the pathogen cannot survive there or has been eradicated (structural zeros). Other locations harbor the pathogen but a given survey fails to detect it during that sampling round (sampling zeros) — e.g., low prevalence or imperfect tests.
 - **Access to specialist treatments / prescriptions**: Some patients will never be eligible for a specific specialist-only procedure or advanced drug (structural zeros) — think of treatments restricted by age, comorbidity, or insurance coverage. Other patients are eligible but simply don't receive the treatment during the study window (sampling zeros) because they don't yet need it, they defer care, or the treatment is rarely used in short follow-up.

```{mermaid}
flowchart LR
  A["Healthcare counts<br/>(e.g. clinic visits)\nwith many zeros"] --> B{Why are there zeros?}
  B --> C["Structural zeros\n(can never have events)"]
  B --> D["Sampling zeros\n(could have events)"]

  C --> C1["Remote residents\n(no clinic access)"]
  C --> C2["Excluded by policy\nor eligibility criteria"]

  D --> D1["Healthy people\n(no care needed during window)"]
  D --> D2["Random non-attendance\n(or short study window)"]

  style C fill:#f8d7da,stroke:#c82333,stroke-width:1px
  style D fill:#d1ecf1,stroke:#0dcaf0,stroke-width:1px
```

_Figure: In healthcare data, zeros can come from fundamentally different processes — policy/access barriers (structural) versus not needing or not attending care during the observation window (sampling). Modeling these separately is what zero-inflated models enable._

## Theoretical Foundation: Generalized Linear Models

### Why GLMs Matter for Zero-Inflated Models

Before diving into zero-inflated models, we need to understand **Generalized Linear Models (GLMs)** — the statistical framework that makes these advanced models possible. Think of GLMs as a powerful extension of ordinary linear regression that can handle different types of data: counts, yes/no responses, continuous positive values, and more.

**The key insight**: Regular linear regression assumes your outcome is continuous and normally distributed (like height or temperature). But what if you're counting discrete events (number of hospital visits) or dealing with yes/no outcomes (did the patient develop diabetes)? GLMs provide the mathematical machinery to handle these situations properly.

Zero-inflated models are sophisticated GLMs that combine two simpler GLM components — that's why understanding GLMs is essential.

### The Exponential Family: The Mathematical Foundation

The **exponential family** is a collection of probability distributions that share a common mathematical structure. This might sound abstract, but it includes most distributions you encounter in practice: normal, Poisson, binomial, gamma, and many others.

**Why does this matter?** Distributions in the exponential family have nice mathematical properties that make them:

- Computationally efficient to work with
- Theoretically well-understood  
- Capable of being combined into complex models like zero-inflated models

A distribution belongs to the exponential family if its probability density can be written as:

$$f(y|\theta, \phi) = \exp\left\{\frac{y\theta - b(\theta)}{\phi} + c(y, \phi)\right\}$$

**You don't need to memorize this!** The only important part is that this format essentially *standardises* how we work with different distributions. The components are:

- $\theta$ is the **natural parameter** (relates to the mean)
- $\phi$ is the **dispersion parameter** (relates to the variance)
- $b(\cdot)$ and $c(\cdot)$ are specific functions that define each distribution type

#### Key Examples with Real-World Context

**Bernoulli Distribution** (for yes/no outcomes):
$$f(y|\mu) = \mu^y(1-\mu)^{1-y} = \exp\left\{y \log\left(\frac{\mu}{1-\mu}\right) + \log(1-\mu)\right\}$$

**Real-world use**: Did the patient develop diabetes? Did the treatment work? Will the customer purchase?

This gives us:

- Natural parameter: $\theta = \log\left(\frac{\mu}{1-\mu}\right)$ (logit transformation)
- Canonical link: logit link

**Why logit link?** Probabilities must be between 0 and 1, but linear combinations can be any value. The logit link maps any real number to a valid probability.

**Practical Healthcare Example**: Predicting whether a patient will develop Type 2 diabetes based on BMI, age, and family history.

- **The setup**: Bernoulli distribution with logit link $\log\left(\frac{\pi}{1-\pi}\right) = X\beta$
- **Interpretation**: $\exp(\beta_j)$ = odds ratio for a unit change in predictor $x_j$
- **In plain English**: If BMI has coefficient 0.1, then each unit increase in BMI multiplies the odds of diabetes by $e^{0.1} = 1.105$ (about 10.5% increase)
- **Key benefit**: The logit link ensures our predicted probabilities stay between 0 and 1

**Poisson Distribution** (for counting rare events):
$$f(y|\mu) = \frac{\mu^y e^{-\mu}}{y!} = \exp\{y \log \mu - \mu - \log y!\}$$

**Real-world use**: Counting hospital admissions per month, number of accidents per day, or doctor visits per year.

The Poisson gives us:

- Natural parameter: $\theta = \log \mu$ (working on the log scale)
- Cumulant function: $b(\theta) = e^\theta = \mu$ 
- Canonical link: $\log(\mu)$ (log link)

**Why log link?** Because counts can't be negative, but our predictors might push the linear combination negative. The log link ensures our predicted counts are always positive.

**Practical Healthcare Example**: Modeling the number of emergency department visits per patient per year based on age, chronic conditions, and socioeconomic status.

- **The setup**: Poisson distribution with log link $\log(\lambda) = X\beta$  
- **Interpretation**: $\exp(\beta_j)$ = multiplicative effect on the rate for a unit change in $x_j$
- **In plain English**: If having diabetes has coefficient 0.5, then diabetic patients have $e^{0.5} = 1.65$ times the visit rate (65% more visits)
- **Key assumption**: $E[Y] = \text{Var}(Y) = \lambda$ (equidispersion)
- **When this breaks down**: If you see many more zeros than Poisson predicts, or if the variance is much larger than the mean, you need zero-inflated or overdispersed models.

### Generalized Linear Models (GLMs): Putting It Together

Think of a GLM as having three essential ingredients, like a recipe:

1. **Linear predictor** ($\eta = X\beta$): This is the familiar part — a linear combination of your predictors
2. **Link function** ($g(\mu) = \eta$): This connects your linear predictor to the expected value of your outcome
3. **Response distribution**: From the exponential family (Poisson, binomial, etc.)

```{mermaid}
flowchart LR
    A["Predictors<br/>(age, treatment, etc.)"] --> B["Linear Combination<br/>η = β₀ + β₁X₁ + β₂X₂ + ..."]
    B --> C["Link Function<br/>g(μ) = η"]
    C --> D["Expected Response<br/>μ (mean count, probability, etc.)"]
    D --> E["Actual Observations<br/>Y ~ Distribution(μ)"]
    
    style A fill:#e1f5fe
    style B fill:#f3e5f5
    style C fill:#fff3e0
    style D fill:#e8f5e8
    style E fill:#fff8e1
```

**The beauty of this system**: You can mix and match different link functions and distributions to model different types of data appropriately.

## Practical Implementation: Foundation Models

### Package Installation

Before starting, install the required packages:

```{r install}
#| eval: false
# Install packages if not already available
install.packages(c("glmmTMB", "DHARMa", "performance", "parameters", "MASS", 
                   "ggplot2", "dplyr", "gridExtra", "knitr", "caret", "catdata"))
```

### Loading Healthcare Data

We'll use a real healthcare dataset to demonstrate zero-inflated models. The `medcare` dataset from the `catdata` package contains information about healthcare utilization among Medicare beneficiaries.

```{r load_data}
#| message: false
#| warning: false
# Load required packages
library(glmmTMB)
library(DHARMa)
library(performance)
library(parameters)
library(MASS)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(knitr)
library(catdata)

# Load the medical care dataset
data(medcare)

# Examine the structure of the data
head(medcare)
glimpse(medcare)
```

**What this dataset contains:**

- `ofp`: Number of physician office visits (our count outcome)
- `hosp`: Number of hospital stays
- `healthpoor`: Individual has poor health (reference: average health)
- `healthexcellent`: Individual has excellent health  
- `numchron`: Number of chronic conditions
- `male`: Gender (female = 0, male = 1)
- `age`: Age (centered around 60)
- `married`: Marital status (married = 1, else = 0)
- `school`: Years of education

### Understanding the Data: Why Zero-Inflation Matters

Let's explore the key characteristics that suggest we need zero-inflated models:

```{r data_exploration}
#| echo: false
#| code-fold: true
#| code-summary: "Show data exploration code"
# Basic statistics for doctor visits — print as a compact table
library(dplyr)
library(knitr)

stats_tbl <- tibble::tibble(
  Metric = c("Proportion of zeros", "Mean visits", "Variance", "Variance-to-mean ratio"),
  Value = c(
    round(mean(medcare$ofp == 0), 3),
    round(mean(medcare$ofp), 2),
    round(var(medcare$ofp), 2),
    round(var(medcare$ofp)/mean(medcare$ofp), 2)
  )
)

knitr::kable(stats_tbl, caption = "Doctor visits (ofp) summary")

# First 10 frequency counts
freqs <- as.data.frame(table(medcare$ofp))
names(freqs) <- c("visits", "count")
knitr::kable(head(freqs, 10), caption = "First 10 visit counts")
```

**Key observations:**

- **High proportion of zeros**: Many people didn't visit a doctor at all
- **Overdispersion**: Variance much larger than mean (typical of real healthcare data)
- **Right-skewed distribution**: Most people have few visits, but some have many

### Visualizing Zero-Inflation

```{r visualization}
#| echo: false
#| code-fold: true
#| code-summary: "Show visualization code"
# Create comprehensive visualization
p1 <- ggplot(medcare, aes(x = ofp)) +
  geom_histogram(binwidth = 1, fill = "lightblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Doctor Visits", 
       x = "Number of Office Visits", y = "Frequency") +
  theme_minimal() +
  xlim(0, 20)  # Focus on reasonable range

# Zero inflation by health status  
p2 <- medcare %>%
  mutate(health_status = case_when(
    healthpoor == 1 ~ "Poor",
    healthexcellent == 1 ~ "Excellent", 
    TRUE ~ "Average"
  )) %>%
  # ensure desired order for plotting
  mutate(health_status = factor(health_status, levels = c("Poor", "Average", "Excellent"))) %>%
  group_by(health_status) %>%
  summarise(prop_zeros = mean(ofp == 0),
            mean_visits = mean(ofp), 
            .groups = 'drop') %>%
  ggplot(aes(x = health_status, y = prop_zeros)) +
  geom_col(fill = "coral", alpha = 0.7, color = "black") +
  labs(title = "Proportion of Zero Visits by Health Status", 
       x = "Health Status", y = "Proportion of Zeros") +
  theme_minimal()

# Zero inflation by chronic conditions
p3 <- medcare %>%
  group_by(numchron) %>%
  summarise(prop_zeros = mean(ofp == 0),
            count = n(),
            .groups = 'drop') %>%
  filter(count >= 20) %>%  # Only groups with sufficient data
  ggplot(aes(x = factor(numchron), y = prop_zeros)) +
  geom_col(fill = "lightgreen", alpha = 0.7, color = "black") +
  labs(title = "Proportion of Zero Visits by Chronic Conditions", 
       x = "Number of Chronic Conditions", y = "Proportion of Zeros") +
  theme_minimal()

# Combine plots
print(p1)
print(p2)
print(p3)
```

**What we observe:**

- People in excellent health have many more zero visits (structural zeros)
- Those with more chronic conditions rarely have zero visits
- This suggests **two distinct processes** generating zeros

### Foundation Models: Step-by-Step Analysis

Now let's build our understanding by starting with simpler models before moving to zero-inflated approaches. This step-by-step approach helps us understand:

1. **The zero-inflation component**: By modeling zeros vs non-zeros, we identify factors that create "structural zeros"
2. **The count component**: By modeling visit counts, we understand factors affecting healthcare utilization intensity  
3. **The need for complex models**: By comparing simple models to our data, we see why zero-inflated models are necessary

This methodical approach mirrors the actual components of zero-inflated models, making the final models more interpretable.

#### Step 1: Logistic Regression for Zero vs Non-Zero

First, let's model the probability of having **any** doctor visits versus none. This analysis mirrors the zero-inflation component of ZIP/ZINB models.

**Why this matters**: Understanding what predicts zero visits helps us identify "structural zeros" - people who systematically don't use healthcare due to access barriers, excellent health, or philosophical reasons, rather than random chance.

```{r logistic_analysis}
#| code-fold: true
#| code-summary: "Show logistic regression analysis"
# Create binary indicator: 1 = zero visits, 0 = any visits
medcare$no_visits <- ifelse(medcare$ofp == 0, 1, 0)

# Fit logistic regression
logistic_model <- glm(no_visits ~ healthpoor + healthexcellent + numchron + age + male + married, 
                     data = medcare, family = binomial())

summary(logistic_model)

# Interpret key coefficients using easystats
library(parameters)

# Get model parameters with confidence intervals and interpretation
model_params <- model_parameters(logistic_model, exponentiate = TRUE)
print(model_params)

# Create interpretation summary
coefs <- coef(logistic_model)
or_values <- exp(coefs)

interpretation_df <- data.frame(
  Variable = c("Health (excellent vs average)", "Health (poor vs average)", 
               "Each chronic condition", "Age (per year)"),
  OR = round(c(or_values["healthexcellent"], or_values["healthpoor"], 
               or_values["numchron"], or_values["age"]), 2),
  Interpretation = c(
    paste0(round((or_values["healthexcellent"] - 1) * 100, 1), "% higher odds of zero visits"),
    paste0(round((or_values["healthpoor"] - 1) * 100, 1), "% higher odds of zero visits"),
    paste0(round((1 - or_values["numchron"]) * 100, 1), "% lower odds per condition"),
    paste0(round((1 - or_values["age"]) * 100, 1), "% lower odds per year")
  )
)

print(interpretation_df, row.names = FALSE)
```

**What this tells us:**

- **Excellent health**: Much higher odds of zero visits (structural zeros)
- **Poor health**: Higher odds of zero visits (potentially due to access barriers or avoidance)
- **Chronic conditions**: Dramatically reduce odds of zero visits
- **Age**: Older patients less likely to have zero visits

#### Step 2: Poisson Regression for Count Data

Now let's model the count of visits using standard Poisson regression. This analysis mirrors the count component of ZIP/ZINB models.

**Why this approach**: Poisson regression is the foundation for count data modeling. By starting here, we can:

- Understand the baseline relationships between predictors and visit counts
- Identify overdispersion (variance > mean), which suggests we need more flexible models
- Establish a baseline for model comparison

**Key Poisson assumptions:**

1. **Independence**: Each visit is independent 
2. **Equidispersion**: Variance equals the mean ($\text{Var}(Y) = E[Y]$)
3. **Non-negative integers**: Counts are 0, 1, 2, 3, etc.

When these assumptions are violated (especially #2), we need more sophisticated models.

```{r poisson_analysis}
#| code-fold: true
#| code-summary: "Show Poisson regression analysis"
# Fit Poisson model
poisson_model <- glm(ofp ~ healthpoor + healthexcellent + numchron + age + male + married, 
                    data = medcare, family = poisson())

summary(poisson_model)

# Check for overdispersion using performance package
overdispersion_check <- check_overdispersion(poisson_model)
print("=== OVERDISPERSION CHECK ===")
print(overdispersion_check)

# Additional performance diagnostics
model_performance_metrics <- model_performance(poisson_model)
print(model_performance_metrics)

# Interpret key coefficients using easystats
poisson_params <- model_parameters(poisson_model, exponentiate = TRUE)
print("=== KEY INTERPRETATIONS (Multiplicative Effects) ===")
print(poisson_params)
```

**Critical insight**: The massive overdispersion (ratio >> 1) tells us that:

- Standard Poisson is **inadequate** for this data
- We need models that account for both **excess zeros** and **overdispersion**
- This is exactly what zero-inflated negative binomial models address!

**Why overdispersion matters**: In Poisson regression, we assume the variance equals the mean. When the variance is much larger (overdispersion), our standard errors are too small, leading to overconfident statistical tests and potentially incorrect conclusions about which variables are significant.

#### Step 3: Comparing Observed vs Expected Zeros

Let's see how many zeros each model predicts compared to what we actually observe. This is the crucial diagnostic that tells us whether we need zero-inflated models.

**The logic**: If Poisson regression adequately captures the data-generating process, it should predict approximately the right number of zeros. If we observe many more zeros than Poisson predicts, this suggests "excess zeros" that require zero-inflated modeling.

**Mathematical background**: For a Poisson distribution with parameter $\lambda$, the probability of observing zero is $P(Y=0) = e^{-\lambda}$. We can use this to calculate expected zeros from our fitted model.

```{r zero_comparison}
#| code-fold: true
#| code-summary: "Show zero-inflation comparison analysis"
# Observed zeros
observed_zeros <- sum(medcare$ofp == 0)
total_obs <- nrow(medcare)

# Predicted zeros from Poisson model
poisson_pred <- predict(poisson_model, type = "response")
poisson_zero_prob <- exp(-poisson_pred)  # P(Y=0) for Poisson
expected_zeros_poisson <- sum(poisson_zero_prob)

# Use performance package to check zero-inflation
zero_inflation_check <- check_zeroinflation(poisson_model)
print("=== ZERO INFLATION CHECK (Performance Package) ===")
print(zero_inflation_check)

# Manual comparison for educational purposes
observed_zeros <- sum(medcare$ofp == 0)
total_obs <- nrow(medcare)

# Predicted zeros from Poisson model
poisson_pred <- predict(poisson_model, type = "response")
poisson_zero_prob <- exp(-poisson_pred)  # P(Y=0) for Poisson
expected_zeros_poisson <- sum(poisson_zero_prob)

# Create comparison table
zero_comparison <- data.frame(
  Metric = c("Observed zeros", "Expected zeros (Poisson)", "Excess zeros"),
  Count = c(observed_zeros, round(expected_zeros_poisson), 
            observed_zeros - round(expected_zeros_poisson)),
  Percentage = c(round(100*observed_zeros/total_obs, 1),
                round(100*expected_zeros_poisson/total_obs, 1),
                round(100*(observed_zeros - expected_zeros_poisson)/total_obs, 1))
)

print("=== ZERO INFLATION COMPARISON ===")
print(zero_comparison, row.names = FALSE)

if(observed_zeros > expected_zeros_poisson * 1.2) {
  print("CONCLUSION: Clear evidence of ZERO-INFLATION!")
} else {
  print("CONCLUSION: Minimal zero-inflation")
}
```

**The smoking gun**: We observe far more zeros than Poisson predicts, confirming we need zero-inflated models.

## Zero-Inflated Models: Theory and Implementation

### The Core Concept: Two-Process Thinking

Now that we've established the need for zero-inflated models, let's understand exactly **how** they work. The key insight is deceptively simple: **not all zeros are created equal**.

Think of zero-inflated models as having two interconnected parts working together, like a two-stage screening process:

```{mermaid}
flowchart LR
    A["New Patient"] --> B{"Stage 1: Zero-Inflation Process<br/>Will this person ever<br/>access healthcare?"}
    B -->|"No (π probability)"| C["Structural Zero<br/>(Guaranteed 0 visits)"]
    B -->|"Yes (1-π probability)"| D{"Stage 2: Count Process<br/>How many visits<br/>will they have?"}
    D --> E["Sampling Zero<br/>(Could visit but didn't)"]
    D --> F["1+ visits<br/>(Standard count model)"]
    
    style C fill:#f8d7da,stroke:#c82333,stroke-width:2px
    style E fill:#d1ecf1,stroke:#0dcaf0,stroke-width:1px
    style F fill:#d4edda,stroke:#28a745,stroke-width:1px
```

**The beauty of this approach**: We can model the factors that influence each stage separately, giving us much richer insights than simple count models.

### Real-World Healthcare Example: Breaking Down the Process

Let's apply this two-stage thinking to our medical care data. Imagine we're trying to understand doctor visit patterns:

**Stage 1 - Zero-Inflation Process** ("Will this person access healthcare at all?")

- **Factors pushing toward structural zeros**: Excellent health status, young age, male gender, geographic barriers, insurance issues
- **Question**: What's the probability this person belongs to the "never visits doctors" group?

**Stage 2 - Count Process** ("Given they do access healthcare, how often?")

- **Factors affecting visit frequency**: Number of chronic conditions, poor health status, age, social support
- **Question**: For people who do use healthcare, what drives the number of visits?

This separation allows us to identify, for example, that **excellent health** primarily affects whether someone visits doctors at all (Stage 1), while **chronic conditions** primarily affect how often they visit (Stage 2).

### Mathematical Formulation: Making the Intuition Concrete

Don't let the mathematics intimidate you — they're just formalizing the two-stage process we described above.

#### Zero-Inflated Poisson (ZIP) Model

The ZIP model mathematically expresses our two-stage process:

**For zeros (no doctor visits):**
$$P(Y = 0) = \pi + (1-\pi)e^{-\lambda}$$

**For positive counts (1, 2, 3, ... visits):**
$$P(Y = k) = (1-\pi)\frac{\lambda^k e^{-\lambda}}{k!}, \quad k = 1, 2, 3, ...$$

**Breaking this down in plain English:**

- $\pi$ = probability of being a "structural zero" (never visits doctors)
- $(1-\pi)$ = probability of being a "potential healthcare user"  
- $\lambda$ = expected visit rate for potential users
- $e^{-\lambda}$ = probability a potential user happens to have zero visits (sampling zero)

**The key insight**: Total probability of zero visits = structural zeros + sampling zeros from potential users.

**Practical Healthcare Interpretation:**

Let's say we estimate $\pi = 0.15$ and $\lambda = 3.2$ for a typical patient profile:

- **15%** of patients are structural zeros (never visit doctors)
- **85%** are potential healthcare users with average rate 3.2 visits/year
- Among potential users, $e^{-3.2} = 4.1\%$ happen to have zero visits by chance
- **Total zero probability** = 15% + 85% × 4.1% = 18.5%

#### Understanding Mean and Variance

**Expected visits:** $E[Y] = (1-\pi)\lambda$
- Only the $(1-\pi)$ proportion who are potential users contribute to the mean
- In our example: $E[Y] = 0.85 \times 3.2 = 2.72$ visits on average

**Variance:** $\text{Var}(Y) = \lambda(1-\pi)(1 + \pi\lambda)$
- Always larger than the mean: $\text{Var}(Y) > E[Y]$ (natural overdispersion)
- In our example: $\text{Var}(Y) = 3.2 \times 0.85 \times (1 + 0.15 \times 3.2) = 4.04$

**Why this matters**: The built-in overdispersion (variance > mean) helps explain the extra variability we see in real healthcare data that simple Poisson models miss.

#### Zero-Inflated Negative Binomial (ZINB) Model

The ZINB model adds another layer of flexibility for handling even more overdispersion:

**For zeros:**
$$P(Y = 0) = \pi + (1-\pi)\left(\frac{r}{r+\mu}\right)^r$$

**For positive counts:**
$$P(Y = k) = (1-\pi)\frac{\Gamma(k+r)}{k!\Gamma(r)}\left(\frac{r}{r+\mu}\right)^r\left(\frac{\mu}{r+\mu}\right)^k$$

**What's different from ZIP:**

- Uses negative binomial instead of Poisson for the count component
- Parameter $r$ controls additional overdispersion within the potential user group
- When $r \to \infty$, ZINB reduces to ZIP (nested models)

**When to use ZINB over ZIP:**

- When you have **excess zeros AND** heavy overdispersion in the count component
- Healthcare data often shows both patterns: some people never use services (excess zeros) and among users, visit patterns are highly variable (overdispersion)

### Building Intuition: A Step-by-Step Example

Let's walk through fitting our first zero-inflated model using the healthcare data, building intuition at each step.

#### Step 1: Choosing Variables for Each Component

**Key decision**: Which variables should influence the zero-inflation process vs the count process?

**Theoretical guidance for healthcare data:**

- **Zero-inflation predictors**: Factors affecting healthcare access/utilization barriers
  - Health status (excellent health → less likely to need care)
  - Demographics (young males often avoid preventive care)
  - Socioeconomic factors (insurance, geographic access)

- **Count predictors**: Factors affecting intensity of healthcare use among users
  - Chronic conditions (more conditions → more visits)
  - Poor health status (sicker people visit more often)
  - Age (older patients typically have more complex needs)

**In practice**: You can use the same variables in both components or different sets, depending on your theoretical understanding and empirical evidence.

### Practical Implementation: From Theory to Code

Now let's implement zero-inflated models using our healthcare data, building step-by-step understanding.

#### Setting Up the Healthcare Models

```{r healthcare_zero_inflated}
#| code-fold: true
#| code-summary: "Show zero-inflated model implementation"
# We'll fit both ZIP and ZINB models to our healthcare data
# Using the same structure as our foundation models but with zero-inflation

# Zero-Inflated Poisson (ZIP) model
zip_model <- glmmTMB(ofp ~ healthpoor + healthexcellent + numchron + age + male + married,
                     ziformula = ~ healthexcellent + age + male,
                     data = medcare,
                     family = poisson())

# Zero-Inflated Negative Binomial (ZINB) model  
zinb_model <- glmmTMB(ofp ~ healthpoor + healthexcellent + numchron + age + male + married,
                      ziformula = ~ healthexcellent + age + male,
                      data = medcare,
                      family = nbinom2())

# Display summaries with interpretation
print("=== Zero-Inflated Poisson Model ===")
summary(zip_model)

print("=== Zero-Inflated Negative Binomial Model ===")
summary(zinb_model)
```

**Understanding the Model Syntax:**

- **Main formula**: `ofp ~ healthpoor + healthexcellent + numchron + age + male + married`  
  - Predicts visit **counts** for people who do use healthcare
  - Focuses on factors affecting utilization intensity

- **Zero-inflation formula**: `ziformula = ~ healthexcellent + age + male`  
  - Predicts probability of being a "structural zero" (never visits)
  - Focuses on factors affecting access/participation

**Why this variable selection makes sense:**

- **Excellent health** appears in zero-inflation (affects whether you visit at all) but count model captures this through both excellent and poor health
- **Age and gender** affect both participation and intensity differently  
- **Chronic conditions** primarily affect intensity (people with more conditions visit more), not whether they participate

### Interpreting Zero-Inflated Model Results

Understanding zero-inflated model output requires interpreting **two separate sets of coefficients**. Let's break this down systematically.

```{r zi_interpretation}
#| code-fold: true
#| code-summary: "Show interpretation of zero-inflated model results"
# Extract and interpret results from ZINB model using easystats
zinb_params <- model_parameters(zinb_model)
print("=== Complete Model Parameters ===")
print(zinb_params)

# Separate the two components for clearer interpretation
count_coefs <- fixef(zinb_model)$cond
zi_coefs <- fixef(zinb_model)$zi

print("=== COUNT MODEL COEFFICIENTS (Log Scale) ===")
print(round(count_coefs, 3))
print("\n=== ZERO-INFLATION COEFFICIENTS (Logit Scale) ===") 
print(round(zi_coefs, 3))

# Convert to interpretable scales
print("\n=== PRACTICAL INTERPRETATIONS ===")

print("COUNT MODEL - Effects on visit rates (among healthcare users):")
count_effects <- exp(count_coefs)
count_df <- data.frame(
  Variable = names(count_effects),
  Rate_Ratio = round(count_effects, 2),
  Percent_Change = paste0(round((count_effects - 1) * 100, 1), "%"),
  Interpretation = c(
    "Baseline visit rate",
    "Poor health vs average health",
    "Excellent health vs average health", 
    "Each additional chronic condition",
    "Each additional year of age",
    "Male vs female",
    "Married vs unmarried"
  )
)
print(count_df[, c("Variable", "Rate_Ratio", "Interpretation")], row.names = FALSE)

print("\nZERO-INFLATION MODEL - Probability of structural zero (never visits):")
zi_probs <- plogis(zi_coefs)
zi_odds <- exp(zi_coefs)
zi_df <- data.frame(
  Variable = names(zi_probs),
  Odds_Ratio = round(zi_odds, 2),
  Baseline_Prob = round(zi_probs, 3),
  Interpretation = c(
    "Baseline odds of never visiting",
    "Excellent health effect on never visiting",
    "Age effect on never visiting", 
    "Male effect on never visiting"
  )
)
print(zi_df[, c("Variable", "Odds_Ratio", "Interpretation")], row.names = FALSE)
```

#### Key Healthcare Insights from the Model

**From the Count Component** (visit rates among users):

- **Chronic conditions**: Each additional condition increases visit rate by ~X% (strong effect)
- **Poor health**: Dramatically increases visit frequency among healthcare users
- **Age**: Older patients have more visits when they do use healthcare
- **Excellent health**: Even among healthcare users, excellent health reduces visit frequency

**From the Zero-Inflation Component** (probability of never visiting):

- **Excellent health**: Much higher odds of being a "never visitor"
- **Age**: Older people less likely to be "never visitors" (more health needs emerge)
- **Gender**: Males more likely to be "never visitors" (avoidance of preventive care)

**The complete picture**: These models reveal that excellent health status **primarily** operates by affecting whether people participate in healthcare at all, while chronic conditions **primarily** affect how intensively people use healthcare once they're in the system.

### Comparing Model Performance: Why ZINB Usually Wins

Let's systematically compare our models to understand which performs best and why.

```{r healthcare_model_comparison}
#| code-fold: true
#| code-summary: "Show model comparison analysis"
# Compare all our healthcare models
models <- list(
  "Poisson" = poisson_model,
  "ZIP" = zip_model, 
  "ZINB" = zinb_model
)

# Get comprehensive comparison metrics
comparison_table <- compare_performance(models, metrics = c("AIC", "BIC", "RMSE"))
print("=== MODEL COMPARISON METRICS ===")
print(comparison_table)

# Likelihood ratio tests for nested models
print("\n=== STATISTICAL TESTS FOR MODEL SELECTION ===")

# Test if ZIP is better than Poisson
if("glm" %in% class(poisson_model) && "glmmTMB" %in% class(zip_model)) {
  # For mixed model comparison, use AIC difference
  aic_diff_zip <- AIC(zip_model) - AIC(poisson_model)
  print(paste("ZIP vs Poisson AIC difference:", round(aic_diff_zip, 2)))
  if(aic_diff_zip < -2) {
    print("ZIP significantly better than Poisson")
  }
} else {
  lrt_zip <- anova(zip_model, poisson_model)
  if(!is.null(lrt_zip$`Pr(>Chisq)`)) {
    print(paste("ZIP vs Poisson p-value:", round(lrt_zip$`Pr(>Chisq)`[2], 4)))
  }
}

# Test if ZINB is better than ZIP
aic_diff_zinb <- AIC(zinb_model) - AIC(zip_model)
print(paste("ZINB vs ZIP AIC difference:", round(aic_diff_zinb, 2)))
if(aic_diff_zinb < -2) {
  print("ZINB significantly better than ZIP")
} else {
  print("ZIP and ZINB perform similarly")
}

# Check which model best handles zero-inflation
print("\n=== ZERO INFLATION HANDLING ===")
zi_check_zip <- check_zeroinflation(zip_model)
zi_check_zinb <- check_zeroinflation(zinb_model)
print("ZIP model zero-inflation check:")
print(zi_check_zip)
print("ZINB model zero-inflation check:")
print(zi_check_zinb)
```

**Why ZINB typically wins in healthcare data:**

1. **Addresses both problems**: Excess zeros AND overdispersion in the count component
2. **More flexible**: Can handle the high variability typical in healthcare utilization
3. **Better fit**: Lower AIC/BIC values indicate superior model fit
4. **Robust diagnostics**: Usually passes zero-inflation and overdispersion tests

### Practical Model Building Strategy

Here's a systematic approach to building zero-inflated models:

#### Step 1: Start with Theory

**Before touching the data**, think about:

- What factors might create "structural zeros"?
- What factors affect intensity among participants?
- Should you use the same or different variables in each component?

#### Step 2: Build Progressively

```{r progressive_building}
#| eval: false
#| code-fold: true
#| code-summary: "Show progressive model building strategy"
# Start simple, build complexity
model1 <- glmmTMB(ofp ~ healthexcellent + numchron,  # Basic count model
                  data = medcare, family = poisson())

model2 <- glmmTMB(ofp ~ healthexcellent + numchron,  # Add zero-inflation
                  ziformula = ~ healthexcellent,
                  data = medcare, family = poisson())

model3 <- glmmTMB(ofp ~ healthpoor + healthexcellent + numchron + age,  # More predictors
                  ziformula = ~ healthexcellent + age,
                  data = medcare, family = poisson())

model4 <- glmmTMB(ofp ~ healthpoor + healthexcellent + numchron + age,  # ZINB for overdispersion
                  ziformula = ~ healthexcellent + age,
                  data = medcare, family = nbinom2())

# Compare at each step
compare_performance(model1, model2, model3, model4)
```

#### Step 3: Validate Your Choices

**Ask these questions:**

- Do the coefficient signs make theoretical sense?
- Does the model pass diagnostic tests?
- Are standard errors reasonable (not too large)?
- Does cross-validation support your model choice?

**Next section**: We'll cover comprehensive model diagnostics to ensure our zero-inflated models are trustworthy.

## Model Diagnostics with DHARMa and Performance

Proper diagnostics are crucial for zero-inflated models due to their complex structure.

```{r diagnostics}
#| eval: false
#| fig-height: 8
#| code-fold: true
#| code-summary: "Show DHARMa diagnostic analysis"
# DHARMa diagnostics for ZINB model
residuals_sim <- simulateResiduals(zinb_model, plot = FALSE)
plot(residuals_sim, main = "DHARMa Residuals for ZINB Model")

# Specific diagnostic tests
print("=== DHARMa Diagnostic Tests ===")
zero_test <- testZeroInflation(residuals_sim)
print(paste("Zero-inflation test p-value:", round(zero_test$p.value, 4)))

disp_test <- testDispersion(residuals_sim)  
print(paste("Dispersion test p-value:", round(disp_test$p.value, 4)))

outlier_test <- testOutliers(residuals_sim)
print(paste("Outliers test p-value:", round(outlier_test$p.value, 4)))
```

```{r performance_checks}
#| eval: false
#| code-fold: true
#| code-summary: "Show performance package diagnostics"
# Performance package diagnostics
print("=== Performance Package Diagnostics ===")
model_perf <- model_performance(zinb_model)
print(model_perf)

# Check specific issues
print("Overdispersion check:")
overdisp_check <- check_overdispersion(zinb_model)
print(overdisp_check)

print("Zero-inflation check:")
zi_check <- check_zeroinflation(zinb_model)
print(zi_check)
```

### Model Comparison

```{r model_comparison}
#| eval: false
#| code-fold: true
#| code-summary: "Show complete model comparison"
# Compare different model approaches
models <- list(
  "Poisson" = poisson_model,
  "ZIP" = zip_model, 
  "ZINB" = zinb_model
)

# Extract model comparison metrics
comparison_table <- compare_performance(models, metrics = c("AIC", "BIC", "RMSE"))
print(comparison_table)

# Likelihood ratio tests for nested models
print("=== Likelihood Ratio Tests ===")
# Test ZIP vs Poisson
lrt_zip <- anova(zip_model, poisson_model)
print(paste("ZIP vs Poisson: p-value =", round(lrt_zip$`Pr(>Chisq)`[2], 4)))

# Test ZINB vs ZIP  
lrt_zinb <- anova(zinb_model, zip_model)
print(paste("ZINB vs ZIP: p-value =", round(lrt_zinb$`Pr(>Chisq)`[2], 4)))
```

The ZINB model wins on all criteria, confirming our need for both zero-inflation and overdispersion modeling.

# Advanced topics

This section covers advanced model choices and extensions beyond the core material (hurdle models, alternative count distributions, model selection strategies, prediction and validation techniques).

## Advanced Topics: Hurdle Models

### Conceptual Difference from Zero-Inflated Models

**Zero-Inflated Models**: Zeros arise from a **mixture** of two processes
- Some zeros are "structural" (never could have non-zero)  
- Some zeros are "sampling" (could have been non-zero but weren't)

**Hurdle Models**: **All** zeros arise from one process, positive values from another
- First process: Will there be any events? (Binary)
- Second process: How many events? (Truncated count)

```{mermaid}
graph TD
    A[Observed Zero] --> B{Model Type}
    B --> C[Zero-Inflated: Could be structural OR sampling zero]
    B --> D[Hurdle: Single process determines all zeros]
    
    E[Observed Positive Count] --> F{Model Type}  
    F --> G[Zero-Inflated: Standard count distribution]
    F --> H[Hurdle: Truncated count distribution]
```

### Mathematical Formulation

**Hurdle Model:**
$$P(Y = 0) = \theta$$
$$P(Y = k) = (1-\theta) \times \frac{f_{\text{truncated}}(k|\lambda)}{1-P_{\text{count}}(0)}, \quad k = 1, 2, 3, ...$$

**Key difference**: Hurdle models can handle **zero deflation** (fewer zeros than expected), while zero-inflated models cannot.

### Implementation

```{r hurdle_models}
#| eval: false
#| code-fold: true
#| code-summary: "Show hurdle model implementation"
# Hurdle Poisson model
hurdle_pois <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                       bait_quality + weather_good,
                       ziformula = ~ experience + weather_good,
                       data = fishing_data,
                       family = truncated_poisson())

# Hurdle Negative Binomial model
hurdle_nb <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                     bait_quality + weather_good,
                     ziformula = ~ experience + weather_good,
                     data = fishing_data,
                     family = truncated_nbinom2())

# Compare all models
all_models <- list(
  "Poisson" = poisson_model,
  "ZIP" = zip_model,
  "ZINB" = zinb_model, 
  "Hurdle_Poisson" = hurdle_pois,
  "Hurdle_NB" = hurdle_nb
)

comparison_all <- compare_performance(all_models, metrics = c("AIC", "BIC"))
print(comparison_all)
```

## When to Choose Hurdle vs Zero-Inflated?

**Choose Zero-Inflated when:**
- Strong theoretical justification for two distinct zero-generating processes
- Zeros are genuinely "extra" beyond what the count distribution predicts
- You want to model factors affecting structural vs sampling zeros differently

**Choose Hurdle when:**
- All zeros conceptually arise from the same process  
- You need flexibility to handle zero deflation
- You want to model "participation" vs "intensity" separately

## Advanced Count Distributions

### Alternative Distributions for Count Data

#### Negative Binomial Parameterizations

```{r nb_parameterizations}
#| eval: false
#| code-fold: true
#| code-summary: "Show negative binomial parameterization options"
# NB1: Variance = μ(1 + α) [linear variance]
zinb1_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                       bait_quality + weather_good,
                       ziformula = ~ experience + weather_good,
                       data = fishing_data,
                       family = nbinom1())

# NB2: Variance = μ(1 + μ/k) [quadratic variance] 
zinb2_model <- zinb_model  # This is our original model

print("NB1 vs NB2 comparison:")
nb_comparison <- compare_performance(zinb1_model, zinb2_model)
print(nb_comparison)
```

**When to use NB1 vs NB2:**
- **NB1**: When variance increases linearly with mean
- **NB2**: When variance increases quadratically with mean (more common)

#### Generalized Poisson for Under/Overdispersion

```{r generalized_poisson}
#| eval: false
#| code-fold: true
#| code-summary: "Show generalized Poisson model"
# Generalized Poisson can handle both under and overdispersion
zigp_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                      bait_quality + weather_good,
                      ziformula = ~ experience + weather_good,
                      data = fishing_data,
                      family = genpois())

summary(zigp_model)
```

#### Conway-Maxwell-Poisson for Flexible Dispersion

```{r conway_maxwell}
#| eval: false
#| code-fold: true
#| code-summary: "Show Conway-Maxwell-Poisson model"
# Conway-Maxwell-Poisson provides very flexible dispersion modeling
# Note: Requires additional computational resources
zicomp_model <- glmmTMB(fish_caught ~ group_size + hours_fishing + experience + 
                        bait_quality + weather_good,
                        ziformula = ~ experience + weather_good,
                        data = fishing_data,
                        family = compois())
```

#### Zero-Altered Models for Continuous Data

For continuous data with zeros (e.g., biomass, rainfall):

```{r zero_altered_gamma}
#| eval: false
#| code-fold: true
#| code-summary: "Show zero-altered gamma model for continuous data"
# Example: Zero-Altered Gamma for biomass data
# Create example biomass data
biomass_data <- data.frame(
  treatment = factor(rep(c("control", "fertilized"), each = 200)),
  soil_moisture = rnorm(400, 50, 15),
  temperature = rnorm(400, 25, 5)
)

# Generate zero-altered gamma data
prob_zero <- plogis(-2 + 0.5 * (biomass_data$treatment == "control"))
has_biomass <- rbinom(400, 1, 1 - prob_zero)
gamma_rate <- exp(-1 + 0.3 * (biomass_data$treatment == "fertilized"))
biomass_data$biomass <- ifelse(has_biomass == 0, 0, 
                              rgamma(400, shape = 2, rate = gamma_rate))

# Zero-Altered Gamma model
zag_model <- glmmTMB(biomass ~ treatment + soil_moisture + temperature,
                     ziformula = ~ treatment + soil_moisture,
                     data = biomass_data,
                     family = Gamma(link = "log"))
```


## Predictions and Interpretation

### Making Predictions from Zero-Inflated Models

```{r predictions}
#| eval: false
#| code-fold: true
#| code-summary: "Show prediction examples"
# Create prediction scenarios
new_scenarios <- data.frame(
  group_size = c(2, 4, 2, 4),
  hours_fishing = c(2, 6, 2, 6),
  experience = factor(c("novice", "novice", "expert", "expert"), 
                     levels = c("novice", "intermediate", "expert")),
  bait_quality = c(0, 0, 1, 1),
  weather_good = c(1, 1, 1, 1)
)

# Get our best model (ZINB)
final_model <- zinb_model

# Different types of predictions
predictions <- data.frame(
  new_scenarios,
  expected_catch = predict(final_model, newdata = new_scenarios, type = "response"),
  prob_zero_inflation = predict(final_model, newdata = new_scenarios, type = "zprob"),
  conditional_catch = predict(final_model, newdata = new_scenarios, type = "conditional")
)

predictions$prob_any_fish <- 1 - predictions$prob_zero_inflation

print(predictions)
```

**Interpretation:**
- `expected_catch`: Overall expected number of fish (includes zero-inflation)
- `prob_zero_inflation`: Probability of being a "structural zero" (non-fisher)
- `conditional_catch`: Expected catch IF the person fishes
- `prob_any_fish`: Probability of catching at least one fish

### Marginal Effects Visualization

```{r marginal_effects}
#| eval: false
#| code-fold: true
#| code-summary: "Show marginal effects visualization"
# Create effect data for hours fishing
effect_data <- expand.grid(
  group_size = 3,  # Average group size
  hours_fishing = seq(0, 8, 0.5),
  experience = "intermediate",
  bait_quality = 0,
  weather_good = 1
)

# Calculate predictions with confidence intervals
effect_predictions <- predict(final_model, newdata = effect_data, 
                             type = "response", se.fit = TRUE)

effect_plot_data <- data.frame(
  effect_data,
  predicted = effect_predictions$fit,
  lower = effect_predictions$fit - 1.96 * effect_predictions$se.fit,
  upper = effect_predictions$fit + 1.96 * effect_predictions$se.fit
)

# Plot marginal effect
ggplot(effect_plot_data, aes(x = hours_fishing, y = predicted)) +
  geom_line(color = "blue", size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  labs(title = "Expected Fish Catch by Hours Fishing",
       subtitle = "For intermediate angler, group size 3, good weather",
       x = "Hours Fishing", 
       y = "Expected Fish Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

## Summary and Best Practices

### Key Takeaways

1. **Zero-inflated models are essential** when you have excess zeros from two distinct processes
2. **ZINB often outperforms ZIP** due to its ability to handle additional overdispersion  
3. **Hurdle models offer more flexibility** and can handle zero deflation
4. **Model diagnostics are crucial** - use DHARMa and performance packages
5. **Cross-validation provides robust model selection** beyond just AIC/BIC

### Best Practices Checklist

✅ **Before modeling:**
- Explore your data for excess zeros
- Consider the data generating process theoretically
- Check for overdispersion in standard Poisson models

✅ **During modeling:**
- Compare multiple model types (ZIP, ZINB, Hurdle)
- Use different covariates for count and zero-inflation components when justified
- Check convergence and model diagnostics

✅ **After modeling:**
- Validate with cross-validation or holdout data
- Interpret both components of the model
- Make predictions appropriate to your research questions

### When to Use Each Model Type

| Model Type | Best When | Avoid When |
|------------|-----------|------------|
| **Standard Poisson** | No excess zeros, equidispersed | Overdispersed or zero-inflated |
| **Negative Binomial** | Overdispersed but no excess zeros | Clear zero-inflation |
| **Zero-Inflated Poisson** | Excess zeros, minimal overdispersion | Heavy overdispersion |
| **Zero-Inflated NB** | Excess zeros AND overdispersion | Zero deflation |
| **Hurdle Models** | Want flexibility, possible zero deflation | Strong theory for mixture processes |

### Further Reading

- Zeileis, A., Kleiber, C., & Jackman, S. (2008). Regression models for count data in R. *Journal of Statistical Software*, 27(8), 1-25.
- Lambert, D. (1992). Zero-inflated Poisson regression, with an application to defects in manufacturing. *Technometrics*, 34(1), 1-14.
- Mullahy, J. (1986). Specification and testing of some modified count data models. *Journal of Econometrics*, 33(3), 341-365.

---

*This tutorial provides a comprehensive foundation for zero-inflated modeling. Remember that the choice of model should always be guided by both statistical evidence and theoretical understanding of your data generating process.*
